**Title:** Analyzing and comparing long-term water quality of lakes and reservoirs in Rhode Island and the Northeastern United States using an anomaly approach

### 2020-09-21 manuscript/ecosphere_resubmit/manuscript.Rmd
getting very close on revisions!

---
title: "**Title:** Analyzing and comparing long-term water quality of lakes and reservoirs in Rhode Island and the Northeastern United States using an anomaly approach"
author:
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*3*^ ORISE, Narragansett, RI 02882
- '**Authors:** Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers, S. ^*3*^, Milstead, W. B. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A. ^*2*^'
- '**Corresponding author: hollister.jeff@epa.gov**'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: |
  | **Abstract:** Addressing anthropogenic impacts on aquatic ecosystems is a focus of lake management. Controlling phosphorus and nitrogen can mitigate these impacts, but determining management effectiveness requires long-term datasets. Recent analysis of the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) United States found stable water quality in the northeastern and midwestern United States, however, sub-regional trends may be obscured. We analyze a sub-regional (i.e., 3000 km^2^) trend with the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset. URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years. The LAGOSNE and URIWW datasets allow for comparison of water quality trends at regional and sub-regional scales, respectively. We assess regional (LAGOSNE) and state (URIWW) trends with yearly median anomalies calculated on a per-station basis. Sub-regionally, temperature and chlorophyll *a* increased from 1993 to 2016. Total nitrogen shows a weak increase driven by low years in the early 1990s. Total phosphorus and the nitrogen:phosphorus ratio (N:P) were stable. At the regional scale, the LAGOSNE dataset shows similar trends to prior studies of the LAGOSNE with chlorophyll *a*, total nitrogen, total phosphorus, and N:P all stable over time. In short, algal biomass, as measured by chlorophyll *a* in Rhode Island lakes and reservoirs is increasing, despite stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio. This analysis suggests an association between lake temperature and primary production. Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific anomalies for analyzing for long-term water quality trends. 
  | **Keywords:** Volunteer Monitoring, Citizen Science, Nutrients, Lakes, Trend Analysis, Long Term
---

@@

Previous studies using regional data from the northeastern and midwestern United States (US) have investigated spatial and temporal water quality trends and have shown differences based on scale. Macro-scale (i.e., ~ 1,800,000 km ^2^) drivers of water quality trends are complex and may vary temporally [@lottig2017macroscale]. This complexity can cause nutrient (N and P) trends to have different drivers than ratios of the nutrients [@collins2017lake]. On a regional scale, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary between regions, particularly when comparing the northeastern (i.e. ~400,000 km^2^) and midwestern (i.e. ~1,000,000 km^2^) US regions [@filstrup2014regional; @filstrup2018evidence]. Furthermore, little change in nutrients and chlorophyll *a* was reported over a 25 year period for these regions [@Oliver_2017]. Given what is known about long-term trends in water quality within the broader regions of the northeastern and midwestern United States (US), we were curious if the lack of trends was also present in water quality at a sub-regional scale, using data on the 3,000 km^2^ area that encompasses a number of Rhode Island lakes and reservoirs.

@@

Prior studies have modeled water quality trends across a larger region of the northeastern US that included 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017]. We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by Oliver et al. [-@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017; @sorannolimno2019].  Temperature data were not available, thus we examined trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.  We used the same selection criteria on the LAGOSNE dataset as was applied to the URIWW data. This ensured that both datasets represented the same seasonal time frame.  Exact measurement depth for the LAGOSNE data were not available; however, the water quality data are reported to be either surface or epilimnion samples [@sorannolagosne2017]. At this time, the LAGOSNE data do not extend beyond 2013 thus sites needed to have a sampling event in 1993-2002 and in 2003-2013.

Lastly, to facilitate a comparison of the context of each scale, we use data, also from LAGOSNE, on landscape composition, maximum lake depth, and lake area [@sorannolagosne2017].  We use the 500 meter buffer landscape composition for three classes from the 2011 National Land Cover Dataset: Agriculuture, Forest, and Developed.  The maximum lake depth is from various sources and lake area is calculated from the National Hydropgraphy Dataset waterbody polygons.  Additional details are available in are detailed in Soranno et al. [-@sorannolagosne2017].

@@

*Summarizing site-specific anomalies:* Methods for calculating the site-specific anomalies and the yearly medians are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly median anomalies.", display = "cite")`.  Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].

@@

1. For each site, calculate the annual medians, producing a single median value for each site and year. This step prevents bias from pseudoreplication of multiple measurements of the same site in a given year [@hurlbert1984pseudoreplication].  The per site medians across years are assumed to be independent.
2. Calculate the long-term reference median for each site. This results in a single long-term median for each of the sites.
3. Calculate the anomaly for each annual median at each site by subtracting the annual and reference medians.
4. Summarize by calculating the median anomaly per year for the entire group of sites.  The resultant values are analyzed for a trend over time.

```{r get_year_n, echo=FALSE}
year_num_ww <- figure_data %>%
  filter(source == "uriww") %>%
  select(year, variable, n) %>%
  spread(variable, n)

year_num_lagosne <- figure_data %>%
  filter(source == "lagosne") %>%
  select(year, variable, n) %>%
  spread(variable, n)
```

@@

*Linear regression on annual median anomalies:* Testing for a regression slope being different than zero can be used to test for monotonic trends in water quality data [@helsel2002statistical].  We used these standard procedures to test for positive or negative trends in lake temperature, chlorophyll *a*, TN, TP and TN:TP.  For each parameter, we fit a regression line to the anomalies as a function of year and tested the null hypothesis that no trend existed (e.g., $\beta$~1~ = 0). The slope of this line provides information on the average yearly change of that parameter over the time period studied.

*Assessing regressions for trends:* Traditionally, trends would be determined by assessing "significance" but recent guidelines suggest not using arbitrary p-value cut-offs to assesses significance [@wasserstein2016asa].  Our interpretation of the trends attempts to follow this advice and we assess trends with multiple lines of evidence. For this assessment, we are defining a trend as with the following: 1) p-values are used to determine general levels of statistical support, 2) relative frequency of high and low years in the beginning and end of the time frame suggest an increasing or decreasing patter, and 4) the magnitude of the slope is used to infer an ecologically relevant change.  Using all of this information we determine that a trend exists if it meets at least 2 of the 3 criteria, a weak trend exists if it meets one of the criteria and no trend exists if it meets none of the criteria.

We recognize that ecologically relevant change is system dependent and no single value will be universally appropriate.  We are using several values for this particular assessment, but other values could also be justified.  There are various ways to identify important temperature changes in lakes and reservoirs.  A policy based approach could examine agreements that attempt to limit temperature increases, such as the Paris Agreement, which aims to limit increases over the next century to 2°C, which would be 0.02°C per year and similarly, the state of Rhode Island criteria for temperature increase is 2.2°C or 0.022°C per year over a century [@paris; @ri_wq_regs]. Alternatively, we could look at more ecologically focused temperature changes.  For instance, Winder and Schindler [-@winder2004climate] saw disrupted trophic linkages between phytoplankton and zooplankton with 1.39°C over a 40 year data set, or 0.035°C per year.  For this study, we will use the 0.02°C per year as it would provide protections against other ecologically relevant changes (e.g. 0.035°C per year for trophic linkages), meet local criteria, and would also be indicative of meeting an important policy goal. For changes in chlorophyll, total phosphorus, and total nitrogen an ecologically meaningful trend is one that would result in a change from an oligotrophic state to a mesotrophic state over the course of a century.  We use the values defined by the trophic state limits in Nürnberg [-@nurnberg1996trophic].  For chlorophyll the oligotrophic-mesotrophic limit is 3.5 μg/L and over a century that is 0.035 μg/L per year, for total nitrogen the limit is 350 μg/L, or 3.5 μg/L per year, and the total phosphorus limit is 10 μg/L or 0.1 μg/L per year.  This level for total phosphorus would also protect against As we have defined it, slope magnitudes in excess of these per year values will be considered to have an ecologically meaningful trend.

*Comparison of Rhode Island to the region:* Oliver et al. [-@Oliver_2017] used hierarchical linear models and showed relatively stable water quality in the lakes of the northeastern United States.  While the University of Rhode Island's Watershed Watch data were included in this regional study, we hypothesized that in the case of Rhode Island regional trends were masking sub-regional trends.  Therefore, we decided to reanalyze the LAGOSNE data to compare the trends at the regional scale to the trends at the Rhode Island state scale using the site-specific anomaly and trend analysis approach outlined above.

@@

Lakes and reservoirs in Rhode Island tended to be smaller and shallower, on average than lakes included in the complete LAGOSNE region (`r tab_num("context_summ", display = "cite", caption = "Average landscape context and lake morphometry summary statistics for lakes in URIWW and LAGOSNE.")`).  Also, LAGOSNE lakes tended to occur in landscapes (i.e. a 500 meter buffer) with a higher average percent agriculture, less developed land, and comparable forested area than lakes in Rhode Island (`r tab_num("context_summ", display = "cite", caption = "Average landscape context and lake morphometry summary statistics for lakes in URIWW and LAGOSNE.")`).

@@

On average, the Rhode Island lakes and reservoirs show lower concentrations of nutrients with more nitrogen relative to phosphorus than in the larger LAGOSNE region (`r tab_num("ww_params", display = "cite")`, `r tab_num("lagos_params", display = "cite")`).  Chlorophyll concentrations also show lower average concentrations in Rhode Island (`r tab_num("ww_params", display = "cite")`, `r tab_num("lagos_params", display = "cite")`).  Furthermore, the distribution of chlorophyll based trophic state also shows a larger percentage of oligotrophic and mesotrophic lakes in Rhode Island and more eutrophic and hypereutrophic lakes in the larger LAGOSNE region [`r tab_num("ts_table", display = "cite", caption = "Percentage of lakes and reservoirs in different chlorophyll based trophic states.")`; @nurnberg1996trophic].

@@

Median annual temperature anomalies in lakes and reservoirs appears to be increasing as the slope is greater than the threshold of 0.02 we identified, the p-value suggests some statistical support for a positive trend (slope = `r signif(ww_temp_rl$slope,2)` , p = `r signif(ww_temp_rl$p.value, 2)`), and the majority of years with median temperature greater than the long-term median are occurring in the second half of the time period (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature (median anomaly) in Rhode Island lakes and reservoirs. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles of site-specific anomalies. Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term medians.", display = "cite")`, `r tab_num(name = "trend_summ", caption = "Summary of long term water quality trends in the URIWW and LAGOSENE data.", display = "cite")`).  Chlorophyll *a* is also showing an increasing trend over time (slope = `r signif(ww_chla_rl$slope,2)` , p = `r signif(ww_chla_rl$p.value, 2)`).  The slope of `r signif(ww_chla_rl$slope,2)` μg/L is greater than our threshold of 0.035 μg/L and the above-average years have mostly occurred in the most recent years (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll a (median anomaly).  Panel A. Yearly median chlorophyll a anomalies from the URI Watershed Watch data.  Panel B. Yearly median chlorophyll a anomalies from the LAGOSNE dataset. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles of site-specific anomalies. Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term medians.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).

@@

In general, there was little evidence to suggest broad regional changes in chlorophyll *a* as it showed a very weak positive trend, slight statistical support and above average years spread evenly throughout the time period (slope = `r signif(lagos_chla_rl$slope,2)`, p = `r signif(lagos_chla_rl$p.value, 2)`, `r fig_num(name = "chla_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).  Total nitrogen showed a slight decreasing trend with weak statistical support and less than the 3.5 μg/L that would result in a trophic state change over a century.  The last two years for which LAGOSNE had data on total nitrogen were from a relatively small number of sites (`r tab_num("year_n_lagosne", display = "cite")`). There is little support for a meaningful trend in TN at the regional scale (slope = `r signif(lagos_tn_rl$slope, 2)`, p = `r signif(lagos_tn_rl$p.value, 2)`, `r fig_num(name = "tn_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`). Total phosphorus showed a very small increasing trend and there is statistical support for that trend.  The slope however suggests little support for an ecological meaningful change in total phosphorus (slope = `r signif(lagos_tp_rl$slope, 2)`, p = `r signif(lagos_tp_rl$p.value,2)`, `r fig_num(name = "tp_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`). Lastly, the TN:TP ratio showed little change (slope = `r signif(lagos_np_rl$slope, 2)`, p = `r signif(lagos_np_rl$p.value, 2)`, `r fig_num(name = "np_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).  Although, there appears to be a non-linear trend that suggests increasing TN:TP early in the time period and decreasing TN:TP later, we feel we cannot effectively evaluate this though as total nitrogen measurements were rare in 2011 (`r tab_num("year_n_lagosne", display = "cite")`) and sites that had both nitrogen and phosphorus were even rarer.  This low median anomaly should be interpreted with this caveat in mind.   Taken together we feel our results largely match the findings of Oliver et al. [-@Oliver_2017] that there is stasis in lake nutrients and chlorophyll within the LAGOSNE region. Data for all figures are available as a comma-separated values file, `yearly_average_anomaly.csv` from <https://github.com/usepa/ri_wq_trends> and are archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].

@@

As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs.  However, 2006 and 2009 stand out as not following this trend.  Average May-October air temperature for 2009 was third lowest on record for the 1993 to 2016 time period and also showed the lowest maximum temperatures [@noaatrends].  On the other hand, 2006 was not unusually cool (11th lowest out of 24 years) but it was the wettest year on record for this period [@noaatrends].

>>>>

---
title: "**Title:** Analyzing and comparing long-term water quality of lakes and reservoirs in Rhode Island and the Northeastern United States using an anomaly approach"
author:
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*3*^ ORISE, Narragansett, RI 02882
- '**Authors:** Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers, S. D. ^*3*^, Milstead, W. B. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A. ^*2*^'
- '**Corresponding author: hollister.jeff@epa.gov**'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: |
  | **Abstract:** Addressing anthropogenic impacts on aquatic ecosystems is a focus of lake management. Controlling phosphorus and nitrogen can mitigate these impacts, but determining management effectiveness requires long-term datasets. Recent analysis of the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) United States found stable water quality in the northeastern and midwestern United States, however, sub-regional trends may be obscured. We analyze for a sub-regional (i.e., 3000 km^2^) trend with the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset. URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years. The LAGOSNE and URIWW datasets allow for comparison of water quality trends at regional and sub-regional scales, respectively. We assess regional (LAGOSNE) and sub-regional (URIWW) trends with yearly median anomalies calculated on a per-station basis. Sub-regionally, temperature and chlorophyll *a* increased from 1993 to 2016. Total nitrogen, total phosphorus and the nitrogen:phosphorus ratio (N:P) were stable. At the regional scale, the LAGOSNE dataset shows similar trends to prior studies of the LAGOSNE with chlorophyll *a*, total nitrogen, and N:P all stable over time.  Total phosphorus does show a very slight increase. In short, algal biomass, as measured by chlorophyll *a* in Rhode Island lakes and reservoirs is increasing, despite stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio. Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific anomalies for analyzing for long-term water quality trends. 
  | **Keywords:** Volunteer Monitoring, Citizen Science, Nutrients, Lakes, Trend Analysis, Long Term
---

@@

Previous studies using regional data from the northeastern and midwestern United States (US), approximately 400,000 km^2^ and 1,000,000 km^2^ respectively, have investigated spatial and temporal water quality trends and have demonstrated that drivers explaining changes spatially may not explain changes temporally. These differences may be due to complex interactions occurring at the different scales through time [@lottig2017macroscale]. Additionally, differences between regions can cause nutrient (N and P) trends to have different drivers compared to ratios of the nutrients, and these may or may not align with sub-continental (i.e., ~ 1,800,000 km ^2^) trends [@collins2017lake]. Similarly, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary between regions, particularly when comparing the northeastern and midwestern US regions [@filstrup2014regional; @filstrup2018evidence]. Furthermore, little change in nutrients and chlorophyll *a* was reported over a 25 year period for these regions [@Oliver_2017]. Given what is known about long-term trends in water quality within the broader regions of the northeastern and midwestern United States (US), we were curious if the lack of trends was also present in water quality at a sub-regional scale, using data on the 3,000 km^2^ area that encompasses a number of Rhode Island lakes and reservoirs.

@@

Prior studies have modeled water quality trends across a larger region of the northeastern US within 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017]. We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by Oliver et al. [-@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017; @sorannolimno2019].  Temperature data were not available, thus we examined trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.  We used the same selection criteria on the LAGOSNE dataset as was applied to the URIWW data. This ensured that both datasets represented the same seasonal time frame.  Exact measurement depth for the LAGOSNE data were not available; however, the water quality data are reported to be either surface or epilimnion samples [@sorannolagosne2017]. At this time, the LAGOSNE data do not extend beyond 2013 thus sites needed to have a sampling event in 1993-2002 and in 2003-2013 to be included.

Lastly, we use data, also from LAGOSNE, on landscape composition, maximum lake depth, and lake area [@sorannolagosne2017].  We use the 500 meter buffer landscape composition for three classes from the 2011 National Land Cover Dataset: Agriculture, Forest, and Developed.  The maximum lake depth is from various sources and lake area is calculated from the National Hydropgraphy Dataset waterbody polygons.  Additional details are available in Soranno et al. [-@sorannolagosne2017].

@@

*Summarizing site-specific anomalies:* Methods for calculating the site-specific anomalies and the yearly medians are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly median anomalies.", display = "cite")`.  Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].

@@

1. For each site, calculate the annual medians, producing a single median value for each site and year. This step prevents bias from pseudoreplication of multiple measurements of the same site in a given year [@hurlbert1984pseudoreplication].  The per site medians across years are assumed to be independent.
2. Calculate the long-term reference median for each site. This results in a single long-term median for each of the sites.
3. Calculate the anomaly for each annual median at each site by subtracting the sites reference long-term medians from the sites yearly median.
4. Summarize by calculating the median anomaly per year for the entire group of sites.  The resultant values are analyzed for a trend over time.

```{r get_year_n, echo=FALSE}
year_num_ww <- figure_data %>%
  filter(source == "uriww") %>%
  select(year, variable, n) %>%
  spread(variable, n) %>%
  rename("Year" = "year", "Chlorophyll" = "chla", 
         "Nitrogen:Phosphorus" = "np_ratio", "Temperature" = "temp", 
         "Total Nitrogen" = "total_n", "Total Phosphorus" = "total_p")

year_num_lagosne <- figure_data %>%
  filter(source == "lagosne") %>%
  select(year, variable, n) %>%
  spread(variable, n)  %>%
  rename("Year" = "year", "Chlorophyll" = "chla", 
         "Nitrogen:Phosphorus" = "np_ratio", 
         "Total Nitrogen" = "total_n", "Total Phosphorus" = "total_p")
```

@@

*Linear regression on annual median anomalies:* Testing for a regression slope being different than zero can be used to test for monotonic trends in water quality data [@helsel2002statistical].  We used these standard procedures to look for positive or negative trends in lake temperature, chlorophyll *a*, TN, TP and TN:TP.  For each parameter, we fit a regression line to the anomalies as a function of year and tested the null hypothesis that no trend existed (e.g., $\beta$~1~ = 0).  The slope of this line provides information on the average yearly change of that parameter over the time period studied.

*Assessing regressions for trends:* Traditionally, trends would be determined by assessing "significance" but recent guidelines suggest not using arbitrary p-value cut-offs to assesses significance [@wasserstein2016asa].  Our interpretation of the trends attempts to follow this advice and we assess trends with multiple lines of evidence. For this assessment, we are defining a trend with the following criteria: 1) p-values are used to determine general levels of statistical support, 2) relative frequency of high and low years in the beginning and end of the time frame are used to identify an increasing or decreasing pattern, and 3) the magnitude of the slope is used to infer an ecologically relevant change.  Using all of this information we determine that a trend exists if it meets at least 2 of the 3 criteria, a weak trend exists if it meets one of the criteria and no trend exists if it meets none of the criteria.

We recognize that ecologically relevant change is system dependent and no single value will be universally appropriate.  We are using several values for this particular assessment, but other values could also be justified.  There are various ways to identify important temperature changes in lakes and reservoirs.  A policy based approach could examine agreements that attempt to limit temperature increases, such as the Paris Agreement, which aims to limit increases over the next century to 2°C, which would be 0.02°C per year and similarly, the state of Rhode Island criteria for temperature increase is 2.2°C or 0.022°C per year over a century [@paris; @ri_wq_regs]. Alternatively, we could look at more ecologically focused temperature changes.  For instance, Winder and Schindler [-@winder2004climate] saw disrupted trophic linkages between phytoplankton and zooplankton with 1.39°C over a 40 year data set, or 0.035°C per year.  For this study, we will use the 0.02°C per year as it would provide protections against other ecologically relevant changes (e.g. 0.035°C per year for trophic linkages), meet local criteria, and would also be indicative of meeting an important policy goal.

For changes in chlorophyll, total phosphorus, and total nitrogen we define an ecologically meaningful trend as one that would result in a change from an oligotrophic state to a mesotrophic state over the course of a century.  We use the values defined by the trophic state limits in Nürnberg [-@nurnberg1996trophic].  For chlorophyll the oligotrophic-mesotrophic limit is 3.5 μg/L and over a century that is 0.035 μg/L per year, for total nitrogen the limit is 350 μg/L, or 3.5 μg/L per year, and the total phosphorus limit is 10 μg/L or 0.1 μg/L per year.

```{r np_thresh, include=FALSE, message=FALSE}
d_and_m_1992 <- read_csv("data/downing_mccauley_1992_table_3.csv")
np_thresh <- d_and_m_1992 %>%
  mutate(molar_np = (total_n/total_p) * 2.21) %>%
  group_by(n_lim) %>%
  summarize(median_np = round(median(molar_np),0))
```

An ecologically meaningful trend for TN:TP could be one that would suggest a switch from nitrogen limitation to phosphorus limitation over a century.  To identify this we use information from experimental manipulations of nitrogen in lakes that identify those lakes as being nitrogen limited [@downing1992nitrogen].  In Downing and McCauley [-@downing1992nitrogen], they identified 34 studies and reported whether or not the study found the lake to be nitrogen limited and the ambient total nitrogen and total phosphorus concentration in μg/L.  With this information we calculated a median molar TN:TP ratio for nitrogen limited lakes and for lakes that did not show nitrogen limitation.  The difference between these two values would suggest a possible shift in the limiting nutrients.  The median molar TN:TP for nitrogen limited lakes was `r np_thresh[np_thresh$n_lim == "Yes",]$median_np` and for lakes not limited by nitrogen was `r np_thresh[np_thresh$n_lim == "No",]$median_np`.  The difference between these is `r abs(np_thresh[np_thresh$n_lim == "Yes",]$median_np - np_thresh[np_thresh$n_lim == "No",]$median_np)` and over a century a change of `r abs(np_thresh[np_thresh$n_lim == "Yes",]$median_np - np_thresh[np_thresh$n_lim == "No",]$median_np)/100` per year would be indicative of an ecologically meaningful change.  As we have defined it, slope magnitudes in excess of these per year values (temperature: 0.02, chlorophyll: 0.035, total nitrogen: 3.5, total phosphorus: 0.1, TN:TP: 0.26) will be considered to have an ecologically meaningful trend.

*Comparison of Rhode Island to the region:* Oliver et al. [-@Oliver_2017] used hierarchical linear models and showed relatively stable water quality in the lakes of the northeastern United States.  While the University of Rhode Island's Watershed Watch data were included in this regional study, we hypothesized that, in the case of Rhode Island, regional trends were masking sub-regional trends.  Therefore, we decided to reanalyze the LAGOSNE data to compare the trends at the regional scale to the trends at the Rhode Island state scale using the site-specific anomaly and trend analysis approach outlined above.

@@

Lakes and reservoirs in Rhode Island tended to be smaller and shallower, on average than lakes included in the complete LAGOSNE region (`r tab_num("context_summ", display = "cite", caption = "Average landscape context and lake morphometry summary statistics for lakes in URIWW and LAGOSNE.")`).  Also, Rhode Island lakes tended to occur in landscapes (i.e. a 500 meter buffer) with a lower average percent agriculture, more developed land, and comparable forested area than lakes in the LAGOSNE region (`r tab_num("context_summ", display = "cite", caption = "Average landscape context and lake morphometry summary statistics for lakes in URIWW and LAGOSNE.")`).

@@

On average, the Rhode Island lakes and reservoirs show lower concentrations of nutrients with more nitrogen relative to phosphorus than in the larger LAGOSNE region (`r tab_num("ww_params", display = "cite")`, `r tab_num("lagos_params", display = "cite")`).  Chlorophyll concentrations also show lower average concentrations in Rhode Island (`r tab_num("ww_params", display = "cite")`, `r tab_num("lagos_params", display = "cite")`).  Furthermore, the distribution of chlorophyll based trophic state also shows a larger percentage of oligotrophic and mesotrophic lakes in Rhode Island and more eutrophic and hypereutrophic lakes in the larger LAGOSNE region [`r tab_num("ts_table", display = "cite", caption = "Percentage of lakes and reservoirs in different chlorophyll based trophic states.")`; @nurnberg1996trophic]. These results follow what would be expected given that, on average, LAGOSNE lakes are more dominated by agricultural lands than are Rhode Island lakes (`r tab_num("context_summ", display = "cite", caption = "Average landscape context and lake morphometry summary statistics for lakes in URIWW and LAGOSNE.")`).

@@

Median annual temperature anomalies in lakes and reservoirs appear to be increasing as the slope is greater than the threshold of 0.02 we identified, the p-value suggests some statistical support for a positive trend (slope = `r signif(ww_temp_rl$slope,2)` , p = `r signif(ww_temp_rl$p.value, 2)`), and the majority of years with median temperature greater than the long-term median are occurring in the second half of the time period (`r fig_num(name = "temp_ri", caption = "Twenty year trend for median lake temperature anomaly in Rhode Island. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles. Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.", display = "cite")`, `r tab_num(name = "trend_summ", caption = 'Summary of long term water quality trends in the URIWW and LAGOSENE data. The "++" indicate a positive trend, the "+" indicates a weak positive trend, and the "0" indicates no trend.', display = "cite")`).  Chlorophyll *a* is also showing an increasing trend over time (slope = `r signif(ww_chla_rl$slope,2)` , p = `r signif(ww_chla_rl$p.value, 2)`).  The slope of `r signif(ww_chla_rl$slope,2)` μg/L is greater than our threshold of 0.035 μg/L and the above-average years have mostly occurred in the most recent years (`r fig_num(name = "chla_trend", caption = "Twenty year trend for median chlorphyll anomaly.  Panel A. URI Watershed Watch yearly chlorophyll anomalies.  Panel B. LAGOSNE yearly chlorophyll anomalies. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles. Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).

@@

In general, there was little evidence to suggest broad regional changes in chlorophyll *a* as it showed a very weak positive trend, slight statistical support and above average years spread evenly throughout the time period (slope = `r signif(lagos_chla_rl$slope,2)`, p = `r signif(lagos_chla_rl$p.value, 2)`, `r fig_num(name = "chla_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).  Total nitrogen showed a slight decreasing trend with weak statistical support and less than the 3.5 μg/L that would result in a trophic state change over a century.  The last two years for which LAGOSNE had data on total nitrogen were from a relatively small number of sites (`r tab_num("year_n_lagosne", display = "cite")`). There is little support for a meaningful trend in TN at the regional scale (slope = `r signif(lagos_tn_rl$slope, 2)`, p = `r signif(lagos_tn_rl$p.value, 2)`, `r fig_num(name = "tn_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`). Total phosphorus showed a very small increasing trend and there is statistical support for that trend.  The slope however suggests little support for an ecological meaningful change in total phosphorus (slope = `r signif(lagos_tp_rl$slope, 2)`, p = `r signif(lagos_tp_rl$p.value,2)`, `r fig_num(name = "tp_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`). Lastly, the TN:TP ratio showed little change (slope = `r signif(lagos_np_rl$slope, 2)`, p = `r signif(lagos_np_rl$p.value, 2)`, `r fig_num(name = "np_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`) as none of our criteria for a trend were met.  Although, there appears to be a non-linear trend that suggests increasing TN:TP early in the time period and decreasing TN:TP later, we feel we cannot effectively evaluate this though as total nitrogen measurements were rare in 2011 (`r tab_num("year_n_lagosne", display = "cite")`) and sites that had both nitrogen and phosphorus were even rarer.  This low median anomaly should be interpreted with this caveat in mind.   Taken together we feel our results largely match the findings of Oliver et al. [-@Oliver_2017] that there is stasis in lake nutrients and chlorophyll within the LAGOSNE region. Data for all figures are available as a comma-separated values file, `yearly_average_anomaly.csv` from <https://github.com/usepa/ri_wq_trends> and are archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].

@@

As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs (`r tab_num(name = "trend_summ", display = "cite")`).  However, 2006 and 2009 stand out as not following this trend.  Average May-October air temperature for 2009 was third lowest on record for the 1993 to 2016 time period and also showed the lowest maximum temperatures [@noaatrends].  On the other hand, 2006 was not unusually cool (11th lowest out of 24 years) but it was the wettest year on record for this period [@noaatrends].  These unusual weather patterns may help explain why 2006 and 2009 did not follow the trend.

Trends in total nutrients and the TN:TP ratio are less clear.  While TN showed a weak positive trend, data for the early years (1994-1998) were sparse.  There is no ecological or statistical trend in the years with consistently available data.  The general picture in Rhode Island appears to be no increase in nitrogen, no trend in phosphorus and little to no change in the TN:TP ratio (`r tab_num(name = "trend_summ", display = "cite")`).  We interpret these results as relative stability in nutrients in Rhode Island lakes and reservoirs.
